---
title: "Linear Models"
author:
  - "Ma Tingting"
  - "Trepça Frida"
  - "Unni Narayanan Adhithya"
  - "Valdés Aharon"
  - "Sander Schoepen"
  
date: "`r Sys.Date()`"
output: html_document
---

## Salary gender discrimination
A group of researchers were interested in whether or not there is gender discrimination with respect to employee salaries at a particular firm. To that end a random sample of 132 employees was studied. The researchers collected information on gender, salary and educational level. The final data set Contained the following variables

Gender: A binary variable taking values Female and Male
Salary: Salary perceived in thousands of dollars per year
Education: A binary variable taking values Degree and No degree

#### Asigment 1: 
Refer to the Salary gender discrimination data set. Assume that the sample sizes do not reflect  the importance of the treatment means. Carry out an unbalanced two-way analysis of variance of this data set,  where the response of interest is Salary and the two crossed factors Gender and Education.

#### Asigment 2: 
Refer to the Salary gender discrimination data set. Assume that the sample sizes reflect the importance of the treatment means. Carry out an unbalanced two-way analysis of variance of this data set, where the response of interest is Salary and the two crossed factors Gender and Education.

Compare the results of both analyses and explain their similarities and differences in a formal and intuitive way.
Discuss the appropriateness of the model.

### Note: 
Students are encouraged to look for alternative analysis and  techniques in the literature if models studied in the course are not appropriate to answer the scientific questions. The idea is to mimic the real work of a consulting statistician in the academia or the industry. 

```{r setup, echo=FALSE, include=FALSE}
library(tidyverse)
library(gridExtra)
library(multcomp)
library(car)
library(ggplot2)
library(MASS)
library(lmtest)
library(DescTools)
fol_main <- '/Users/aronvz/Documents/github/kul-lm-as01/'
fol_src  <- '_src/'
fol_out  <- '_out/'
fil_raw  <- 'salary.gender.txt'

setwd(fol_main)

salary.gender.txt <- read.table(paste0(fol_src,fil_raw), 
                                header=TRUE, 
                                skip = 27,
                                quote="\"", sep=",")

summary(salary.gender.txt)
#Salary-Gender-Education 

salary.gender.txt01 <- 
  transform(salary.gender.txt,
            f.gender = as.factor(Gender), 
            f.education = as.factor(Education))

summary(salary.gender.txt01)
#Salary-Gender-Education __ f.gender-f.education

levels(salary.gender.txt01$f.gender)
levels(salary.gender.txt01$f.education)

```

# 2 Checking DataSet
### 2.1 Collecting and organizing data
Collect data from each group and organize it into a suitable format for analysis. Make sure to record the sample size for each group.
```{r pdescr-1, fig.width = 13}
#Exploratory Data Analysis

# Distribution of response variable - Salary
p_ht1 <- ggplot(salary.gender.txt01, aes(Salary)) + 
  geom_histogram(colour = 'black', fill = 'blue', bins=30) +
  ggtitle('Dist. Salary') + 
  xlab('Response - Salary') + 
  theme_bw()

# Distribution of explanatory variable - f.gender
p_br1 <- ggplot(salary.gender.txt01, aes(f.gender, fill = f.gender)) + 
  geom_bar() + 
  ggtitle('Dist. of Gender') + 
  xlab('Explanatory - Gender') + 
  theme_bw()+
  theme(panel.grid = element_blank(), legend.position = 'none')

# Distribution of explanatory variable - f.education
p_br2 <- ggplot(salary.gender.txt01, aes(f.education, fill = f.education)) + 
  geom_bar() + 
  theme(axis.text.x=element_blank())  + 
  ggtitle('Dist. of Education') + 
  xlab('Explanatory - Education') + 
  theme_bw()+
  theme(panel.grid = element_blank(), legend.position = 'none')
grid.arrange(p_br1, p_br2,p_ht1, ncol = 3)

```


### 2.2 Mean of response variable 

```{r pdescr-2, fig.width = 13}
table_unbal_mean <- with(salary.gender.txt01, 
                         tapply(Salary, list(f.gender, f.education), mean))

table_unbal_mean %>% round(2) %>% as.table()

# Frequency table of response variable by gender, education and salary
table_unbal_freq <- with(salary.gender.txt01, 
                         table(f.gender, f.education))
table_unbal_freq

# Data in long format
data_longer <- table_unbal_mean %>% as.data.frame()
data_longer$`f.gender` <- row.names(data_longer)
data_longer <- data_longer %>% 
  pivot_longer(cols = c(Degree, `No degree`),
               names_to = 'degree',
               values_to = 'mean_y')
data_longer

```
### 2.3 Interaction 

```{r pdescr-3, fig.width = 13}
######################################################

#with(salary.gender.txt01, interaction.plot(
#  x.factor = Education, 
#  trace.factor = Gender, 
#  response = Salary,
#  fun = mean, type = "b", legend = T,ylab = "Mean Salary",
#                                main="Interaction plot", pch=c(1,19)))

int1 <- ggplot(data_longer, aes(degree, mean_y)) + 
  geom_line(aes(group = f.gender, color = f.gender), size = 1) + 
  xlab('Degree') + 
  ylab('Mean value of Salary') + 
  ggtitle('Interaction Plot') + 
  theme_bw() + 
  theme(axis.text.y = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.text.x = element_text(size = 13),
        axis.title.x = element_text(size = 13))

int2 <- ggplot(data_longer, aes(f.gender, mean_y)) + 
  geom_line(aes(group = degree, color = degree), size = 1) + 
  xlab('Gender') + 
  ylab('Mean value of Salary') + 
  ggtitle('Interaction Plot') + theme_bw() +   
  theme(axis.text.y = element_text(size = 13),
        axis.title.y = element_text(size = 13),
        axis.text.x = element_text(size = 13),
        axis.title.x = element_text(size = 13))

grid.arrange(int1, int2, ncol = 2)
```

### 2.3 Interaction BoxPlot


```{r pdescr-4, fig.width = 13}
salary.gender.txt01$EdDe <- interaction(
  salary.gender.txt01$Gender,
  salary.gender.txt01$Education)

salary.gender.txt01$EdDe <- ifelse(
  salary.gender.txt01$EdDe=='Male.No degree','M.NoD',
    ifelse(salary.gender.txt01$EdDe=='Male.Degree','M.Deg', 
           ifelse(salary.gender.txt01$EdDe=='Female.No degree','F.NoD','F.Deg'))
) 

bxp1 <- ggplot(salary.gender.txt01, aes(f.gender, y = Salary)) + 
geom_boxplot(outlier.colour="black", outlier.shape=16,
             outlier.size=2, notch=FALSE) +
  xlab('Gender') 

bxp2 <- ggplot(salary.gender.txt01, aes(f.education, y = Salary)) + 
  geom_boxplot(outlier.colour="black", outlier.shape=16,
               outlier.size=2, notch=FALSE) + 
   xlab('Educational Level') +
  theme(axis.text.y=element_blank(),
    axis.title.y.left = element_blank(),
    axis.ticks.y=element_blank())


bxp3 <- ggplot(salary.gender.txt01, aes(EdDe, y = Salary)) + 
  geom_boxplot(outlier.colour="black", outlier.shape=16,
               outlier.size=2, notch=FALSE) + 
  xlab('Educational Level & Gender') + 
  theme(axis.text.y=element_blank(),
    axis.title.y.left = element_blank(),
    axis.ticks.y=element_blank())
 

grid.arrange(bxp1, bxp2, bxp3, ncol = 3)

#boxplot(Salary~Gender, col=0, data = salary.gender.txt01)
#boxplot(Salary~Education, col=0, data=salary.gender.txt01)
#boxplot(Salary~Gender*Education, col=0, data=salary.gender.txt01, 
#        main="Boxplots of Salaries Data")
```

## 3 Assumptions
Check the assumptions: Two-way ANOVA assumes that the data are normally distributed and homoscedastic (i.e., the variances of the groups are equal). Use statistical tests or plots to check whether these assumptions are met.

```{r passum-1, fig.width = 13}

```


## Fit Anova

```{r panova}
salaries.fit<-aov(Salary~Gender*Education,data=salary.gender.txt01)
salaries.fit.summary<-summary(salaries.fit)
salaries.fit.summary
```

The only significant relation is __education__. 

No evidence for an interaction effect.

***** 

comparison using Tukey procedure

```{r ptukey}
salaries.fit.tukey<-TukeyHSD(salaries.fit,which=c("Gender:Education"),conf.level = .95)
plot(salaries.fit.tukey)
salaries.fit.tukey
```

Fit a two-way anova model without interaction effects

```{r twowayanovanointer}
salaries.fit_2<-aov(Salary~Gender+Education,data=salary.gender.txt01)
salaries.fit_2.summary<-summary(salaries.fit_2)
salaries.fit_2.summary
```
### Type I SS

```{r typeIss}

type1.unbalanced1<-aov(Salary~Gender*Education,data = salary.gender.txt01)
type1.unbalanced1.summary<-summary(type1.unbalanced1)
type1.unbalanced1.summary #only Education failed null-hypothesis

type1.unbalanced2<-aov(Salary~Education*Gender,data = salary.gender.txt01)
type1.unbalanced2.summary<-summary(type1.unbalanced2)
type1.unbalanced2.summary #both Education and Gender failed null-hypothesis
```
### Type II SS
```{r typeIIss}
type2.unbalanced<-Anova(lm(Salary~Gender*Education,data = salary.gender.txt01,
                           contrasts=list(Gender="contr.sum", Education="contr.sum")), type="II")
type2.unbalanced #both Education and Gender failed null-hypothesis
```

### Type III SS 
```{r typeIIIss}

type3.unbalanced<-Anova(lm(Salary~Gender*Education,data = salary.gender.txt01,contrasts=list(Gender="contr.sum", Education="contr.sum")), type="III")
type3.unbalanced #both Education and Gender failed null-hypothesis
```
### Decision 
Type I, II or III decision 
  * Type III will be used
  * No significant interaction between Gender and Education.

### Factor Effect Model

```{r definingXmatrix}
data <- salary.gender.txt01
data$Gender <- as.factor(data$Gender)
data$Education <- as.factor(data$Education)
summary(data)
contrasts(data$Gender) = contr.sum
contrasts(data$Education) = contr.sum
model.matrix(~Gender*Education, data)

```
### Creating the indicator variables
```{r indicatorvariables}
data1 <- as.data.frame(model.matrix(~ Gender*Education, data, contrasts = list(Gender="contr.sum",Education="contr.sum"))[,2:4]) 
data1$Salary <-data$Salary
data1

data.full=lm(Salary ~ Gender1+ Education1 +Gender1:Education1, data = data1)
summary(data.full)

data.reduced=lm(Salary ~ Gender1+ Education1, data = data1)
summary(data.reduced)
```


### Test for the interaction
```{r testinteraction}
anova(data.reduced, data.full) #no interaction
```

### analysis of residuals 
```{r analysisofresiduals }
salary.fit <- aov(Salary~Education, data = salary.gender.txt01)
summary(salary.fit) # null-hypothesis rejected

plot(salary.fit)
#plot(res,fitted(salary.fit),
#     xlab="Fitted values",
#     ylab="Studentized residuals",
#     main="Residuals vs fitted values plot")
#abline(h=0,lty="dashed")

plot(salary.gender.txt01$Salary ,rstandard(salary.fit), xlab="Observed values",
     ylab="Studentized residuals",
     main="Residuals vs factor levels plot",axes=F)
axis(side=1)

axis(side=2)
abline(h=0,lty="dashed")

res<-salary.fit$residuals
plot(fitted(salary.fit),res)
abline(0,0)
```
### Studentized residuals

```{r studentizedresiduals }
salary.gender.txt01$rs<-rstandard(salary.fit)
salary.gender.txt01$fit<-fitted.values(salary.fit)
salary.gender.txt01$resi<-salary.fit$residuals
#salary$Education<-as.factor(salary.gender.txt01$Education)
leveneTest(Salary~Education, data=salary.gender.txt01) 
```
has heterosceracicity

### Test for independence
```{r testforindependence }
durbinWatsonTest(salary.fit,alternative="two.sided", 
                 data=salary.gender.txt01) #independent
```

### Normality

```{r normality }
qqnorm(residuals(salary.fit))
qqline(residuals(salary.fit)) #does not look normal
shapiro.test(residuals(salary.fit)) # Passed normality test
ks.test(residuals(salary.fit),"pnorm", alternative="two.sided") 
```

#### Failed normality test
```{r normality2 }
ggplot(salary.gender.txt01, aes(sample=resi))+stat_qq()+stat_qq_line(color = "red")
#ggsave("QQLogPlot.png", path = output_path, dpi = 72,width = 12, height = 12)
ggplot(data = salary.gender.txt01, aes(y = resi, x = Salary)) + 
  geom_point(col = 'blue') + 
  geom_abline(slope = 0)
```

### multiple comparasion
```{r multipleComparasion }
TukInit <- TukeyHSD(salary.fit)
plot(TukInit)
scheffe <- ScheffeTest(salary.fit)
plot(scheffe)
all_pairwise = rbind("Degree-No Degree" = c(-1, 1))
```


### find the best transformation based on boxcox procedure

```{r boxcoxprocedure }

data.boxcox<-boxcox(salary.fit, lambda=seq(-2, 4,by=.1))
lambda.max<-max(data.boxcox$y)
boxcox.lambda <- data.boxcox$x[data.boxcox$y==lambda.max]
boxcox.lambda
salary.gender.txt01$boxcox<-salary.gender.txt01$Salary*boxcox.lambda
socialbox.fit <-aov(boxcox~f.education*f.gender, data = salary.gender.txt01)
summary(socialbox.fit)
leveneTest(boxcox~Education, data=salary.gender.txt01)
bptest(salary.fit)
plot(fitted(salary.fit),resid(salary.fit))
```






